---
title: "NBA Reproducible Data analysis Project"
author: "David Mawa"
date: "04/05/2021"
output:
  html_document: default
---

## Packages:
These are the packages that will be used in this project.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(ggcorrplot)
```

## Introduction:
The National Basketball Association (NBA) is a professional basketball league in North 
America comprising 30 franchises, of which 29 are located in the United States and one in 
Canada (Toronto Raptors). The league was founded in 1946, then named the Basketball 
Association of America (BAA). In 1949, the BAA merged with the rival National Basketball 
League (NBL) to form the National Basketball Association. 
Today the NBA as a whole generates around eight billion U.S. dollars in revenue per season, 
an average per team of around 264 million U.S. dollars in a season(statista.com).
Organized basketball is a game played by five players per team. Historically, these players 
have been assigned to positions defined by the role they play on the court. These positions 
are; the point guard, the shooting guard, the small forward, the power forward and the 
center(wikipedia.org).
The previous season Chicago Bulls was placed 27th out of 30 based on win-loss record.
The purpose of this project is to analysis the given NBA datasets to find the top five starting 
players (one from each position) that the Chicago Bulls team can recruit but also maintaining 
to fill a full team roster. The Chicago Bull's budget for player contracts for the following 
season is $118 million. The analysis will help the Chicago Bulls find quality players for the 
next season with low salaries and help improve the team. The matric I will use to evaluate the 
players will be based on points to minute played ratio (pmp). Calculated as pmp = PTS 
(points)/MP (minute played). 

## Reading the raw datasets:
Uploading the datasets in R.

```{r,message =FALSE, warning==FALSE}
player_salaries <- read_csv("2018-19_nba_player-salaries.csv")
player_stats <- read_csv("2018-19_nba_player-statistics.csv")
team_stats_1 <- read_csv("2018-19_nba_team-statistics_1.csv")
team_stats_2 <- read_csv("2018-19_nba_team-statistics_2.csv")
team_payroll <- read_csv("2019-20_nba_team-payroll.csv")
```


## Datasets description:
This is the brief description of the datasets that will be used in analysis.

2018-19_nba_player-statistics.csv. 
This data file provides total statistics for individual NBA players during the 2018-19 season. Has 708 observations and 29 variables. The variables consist:

player_name : Player Name
Pos :  (PG = point guard, SG = shooting guard, SF = small forward, PF = power forward, C = center) 
Age : Age of Player at the start of February 1st of that season.
Tm : Team
G : Games
GS : Games Started
MP : Minutes Played
FG : Field Goals
FGA : Field Goal Attempts
FG% : Field Goal Percentage
3P : 3-Point Field Goals
3PA : 3-Point Field Goal Attempts
3P% : FG% on 3-Pt FGAs
2P : 2-Point Field Goals
2PA : 2-point Field Goal Attempts
2P% : FG% on 2-Pt FGAs
eFG% : Effective Field Goal Percentage
FT : Free Throws
FTA : Free Throw Attempts
FT% : Free Throw Percentage
ORB : Offensive Rebounds
DRB : Defensive Rebounds
TRB : Total Rebounds
AST : Assists
STL : Steals
BLK : Blocks
TOV : Turnovers
PF : Personal Fouls
PTS : Points

2018-19_nba_player-salaries.csv.
This data file contains the salary for individual players during the 2018-19 NBA season.
Has 576 observations and 3 variables.The variables consist:

player_id : unique player identification number
player_name : player name
salary : year salary in $USD

## Data cleaning.
Looking at the datasets.

```{r, message=FALSE, warning=FALSE}
head(player_salaries)
tail(player_salaries)
dim(player_salaries)

head(player_stats)
tail(player_stats)
dim(player_stats)
```

Missing values handling.
The player salaries data has no missing values.

The player statistics data has 117 missing values.

```{r, message=FALSE, warning=FALSE}
player_salaries %>% summarise(sum(is.na(.)))

player_stats %>% summarise(sum(is.na(.)))
```

Columns that contain the missing values from the player statistics data.

The missing values are from players who played less games. Therefore will drop them.

```{r, message=FALSE, warning=FALSE}
colnames(is.na(player_stats))
colSums(is.na(player_stats)) %>% as.data.frame() 
```

Dropping missing values.

```{r, message=FALSE, warning=FALSE}
player_stats_clean <- player_stats %>% drop_na()
player_stats_clean
```

Merging player salaries data and player statistics clean data.

```{r, message=FALSE, warning=FALSE}
player_data <- full_join(player_salaries, player_stats_clean, by = "player_name")
player_data
```

Duplicates handling.
There are players that are repeated due to season trading.

```{r, message=FALSE, warning=FALSE}
unique(player_data$player_name[duplicated(player_data$player_id)]) %>% as.data.frame()

player_data <- player_data[!duplicated(player_data$player_id), ]
player_data 
```

## Exploratory Data Analysis:

Missing values introduced after the two datasets are merged.
The missing values are mostly from players with no defined position so will also drop them

```{r, message=FALSE, warning=FALSE}
player_data_tidy <- player_data %>% drop_na()
player_data_tidy
```

Salaries Distribution.
From the salary distribution plot, the histogram is skewed to the right.

```{r, message=FALSE, warning=FALSE}
ggplot(data = player_data_tidy, aes(x = salary)) + 
  geom_histogram( color = "black", fill = "blue") + ggtitle("Salary distribution")
```

Salary summary.
The median salary is $3258539
The mean salary is $6967454 

```{r, message=FALSE, warning=FALSE}
summary(player_data_tidy$salary)
```

Questions to answer.
Do players with higher game points have higher salaries?
Do playing positions affects player's salary?

Points vs Salary.
Players that have higher points tend to have high salaries
regardless of the position they player in.

```{r, message=FALSE, warning=FALSE}
ggplot(data = player_data_tidy, aes(x = PTS, y = salary, color = Pos)) +
  geom_point() + ggtitle("player points against salary")
```

Bar plot of positions.
There are a high number of Shooting guard players in 2018-19 
season compared to the other positions. Center position has the least number of players.

```{r, message=FALSE, warning=FALSE}
ggplot(data = player_data_tidy, aes(x = Pos, fill = Pos)) + geom_bar() +
  ggtitle("Number of players for each position in the season 2018-19")
```

Games started by a player against points.
This does not give a proper picture. 
Because there are players that started fewer games but they still have high points.

```{r, message=FALSE, warning=FALSE}
ggplot(data = player_data_tidy, aes(x = GS, y = PTS)) + geom_point() +
  ggtitle("Games started by a player against Points") + geom_smooth(method = lm)
```

Minutes played.
Players that played more minutes have high points.

```{r, message=FALSE, warning=FALSE}
ggplot(data = player_data_tidy, aes(x = MP, y = PTS)) + geom_point() +
  ggtitle("Minutes played by a player vs Points") +geom_smooth(method = lm)
```

Variable correlation.
Salary is correlated with the other variables but not highly.

Points are highly correlated with Turnovers,Steals,Free Throw Attempts, Defensive Rebounds
Free Throws,2-point Field Goal Attempts,2-Point Field Goals,3-Point Field Goal Attempts,
3-Point Field Goals,Field Goals,minutes Played,Games Started,Total Rebounds.

Player's Age is correlated with salary.
High correlation among some of the variables exist as well.


```{r, message=FALSE, warning=FALSE}
cor_data <- player_data_tidy %>% select(-c(player_name, Pos, Tm))

ggcorrplot(cor(cor_data)) +ggtitle("Correlation headmap of the variables")
```

Features extraction.
Extracting the variables to be used for the modeling

```{r, message=FALSE, warning=FALSE}
model_data <- player_data_tidy %>%
  select(c(player_name,salary,Pos,Age,Tm,G,GS,MP,FG,FGA,"3P","3PA","2P","2PA",FT,FTA,DRB,TRB,TOV,PTS))
model_data
```

## Data Modeling:

Simple linear Regression: to test if players points explains salary
Player's point does not explain salary well.
makes a bit of sense since scoring high points are contributed to by other factors. 
forexample, if a player played more minutes, there is a chance he can score more points. 

The model achieved low R-squared of 0.2974.
When points is zero, player salary is $2557795.8. But when points is increased by 1 unit, players salary 
increase by 8708.7. 

```{r, message=FALSE, warning=FALSE}
model1 <- lm(salary~PTS, data = model_data)

summary(model1)
tidy(model1, conf.int = TRUE)
```

Multiple Linear Regression.
Player with better statistics,tend to have high points. 
This model has an R-squared of 0.9979.

```{r, message=FALSE, warning=FALSE}
model_data2 <- player_data_tidy %>%
  select(c(Pos,Age,G,GS,MP,FG,FGA,FT,FTA,DRB,TRB,TOV,PTS))

model2 <- lm(PTS~., data = model_data2)
summary(model2)
tidy(model2, conf.int = TRUE)
```

Finding players for each of the positions.
Conditions used to filter players: a player must atleast have played a minimum of 50 games, 
1000 minute played, salary between 1 Million and 10 Million and minimum points of 1000. 

```{r, message=FALSE, warning=FALSE}
Rec_data <- player_data_tidy %>%
  select(c(player_name,Tm,salary,Pos,G,GS,MP,PTS)) %>% arrange(desc(PTS))
Rec_data
```

Point guard players.

The player with reasonable salary and good pmp for the Point Guard position is 
D'Angelo Russell: salary 7019698, games 81, games started 81, minute played 2448,
points 1712 and points per minute played ratio 0.699. 

Alternative could consider,
Spencer Dinwiddie, salary 1656092, games 68, game started 4, minute played 1914, 
points 1143 and points per minute played ratio 0.597.

```{r, message=FALSE, warning=FALSE}
pos_pg <- Rec_data %>% filter(Pos == "PG") %>% 
  filter(G >=50, MP>=1000, salary>=1000000, salary<=10000000, PTS>=1000) 
pos_pg

#points per minute played(pmp)
pos_pg %>% mutate(pmp = PTS/MP) %>% 
  arrange(desc(pmp))
```

Shooting guard players.

The player to consider for the  position of shooting guard is,
Devin Booker: salary 3314365, games 64, games started 64, minute played 2242,
points 1700 and points to minute played ratio 0.758.

```{r, message=FALSE, warning=FALSE}
pos_sg <- Rec_data %>% filter(Pos == "SG") %>% 
  filter(G >=50, MP>=1000, salary>=1000000, salary<=10000000, PTS>=1000) 
pos_sg
  
#Points per minute played (pmp)
  pos_sg%>% mutate(pmp = PTS/MP) %>% 
  arrange(desc(pmp))
```

Small forward.

The player to consider for the position of small forward is,
Jayson Tatum: salary 6700800, games 79, games started 79, minute played 2455,
points 1243 and points to minute played ratio 0.506. 

```{r, message=FALSE, warning=FALSE}
pos_sf <- Rec_data %>% filter(Pos == "SF") %>% 
    filter(G >=50, MP>=1000, salary>=1000000, salary<=10000000, PTS>=1000) 
  pos_sf
  
#Points per minute played (pgmp)
  pos_sf %>% mutate(pmp = PTS/MP) %>% 
    arrange(desc(pmp))
```

Power forward.

Player to consider for the position of power forward is,
Julius Randle: salary 8641000, games 73, games started 49, minute played 2232  
points 1565 and points per minute played ratio 0.701.

Alternatively we can consider a cheaper option
John Collins: salary 2299080, games 61, games started 59, minute played 1829,
points 1188 and points per minute played ratio 0.650.

```{r, message=FALSE, warning=FALSE}
pos_pf <- Rec_data %>% filter(Pos == "PF") %>% 
    filter(G >=50, MP>=1000, salary>=1000000, salary<=10000000, PTS>=1000) 
  pos_pf
  
#Points per minute played (pgmp)
pos_pf %>% mutate(pmp = PTS/MP) %>% 
    arrange(desc(pmp))
```

Center. 
#Player to consider for the position of center is,
Karl-Anthony Towns: salary 7839435, games 77, games started 77, minute played 2545,
points 1880 and points to minute played ratio 0.739.

```{r, message=FALSE, warning=FALSE}
 pos_c <- Rec_data %>% filter(Pos == "C") %>% 
    filter(G >=50, MP>=1000, salary>=1000000, salary<=10000000, PTS>=1000) 
  pos_c
  
#Points per minute played (pgmp)
  pos_c %>% mutate(pmp = PTS/MP) %>% 
    arrange(desc(pmp))
```


## Player Recommendations.
The top five layers recommended to buy for the next season by Chicago Bulls.
The total cost of these five players could be $33515298.
The remaining $84484702 could be used to find other quality players to fill the team's season roster. 
These top five players have great statistics in terms of points and points per minute played but are
undervalued in the money. Giving them slight better contract could attract them to join the team.
They could be of great benefit to Chicago Bulls in the next NBA season. 

```{r, message=FALSE, warning=FALSE}
recomend<- rbind(pos_c, pos_pf)
recomended <- rbind(recomend, pos_pg)    
recomended_pl <- rbind(recomended,pos_sf)
recomended_players <- rbind(recomended_pl, pos_sg) %>% mutate(pmp = PTS/MP) 
recomended_players[c(1,6,12,19,21),] %>% arrange(desc(pmp))
```

## Summary.
In this project analysis, few hypothesis about the NBA player statistics and salary datasets were
explored. Determined if players with higer game points have higher salary and if playing position 
is a factor that is more correlated to the salary. It was concluded that, there are some other factors 
that are correlated to salary.
It was also found that in the season 2018-19, there were more players in shooting guard position and less
center players. 
The salary distribution is skewed to the right and oints are predicted by other factors. Salary is not
mostly predicted by points. 
Linear and multiple regression models were created that would predict salary and points. 
The analysis has some limitations; removing outliers were not taken into consideration and few selected 
variables were used in the creation of the two models. The metric used could not be considered as 
the best evaluator for selecting the players. Some conditions were also set as a requirement
in selecting the players.

## Reference list.
basketball-reference.com

oopshype.com/salaries

https://www.statista.com/statistics/193442/average-franchise-value-in-the-nba-since-2000/

https://en.wikipedia.org/wiki/Basketball_positions





